---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geotax

<!-- badges: start -->
[![R-CMD-check](https://github.com/alrobles/geotax/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/alrobles/geotax/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

geotax models the probability that two hosts share a pathogen (or any
symbiont) given their phylogenetic or taxonomic distance, following
[Robles-Fernández & Lira-Noriega (2017)](https://doi.org/10.3389/fams.2017.00017).

## Installation

You can install the development version of geotax from [GitHub](https://github.com/alrobles/geotax) with:

``` r
# install.packages("devtools")
devtools::install_github("alrobles/geotax")
```

## Workflow

The v2 API goes from an interaction table plus a host distance matrix
to a fitted host-sharing model:

1. `prepare_taxonomic_tree()` — build a taxonomic tree from a rank table
   (or use any phylogeny, e.g. from [Fish Tree of Life](https://fishtreeoflife.org/)).
2. `prepare_pair_data()` — expand an interaction table into all
   (focal host, target host) pairs per parasite. All known hosts are used
   as focal hosts, so the data set is deterministic.
3. `compare_geotax_slopes()` / `cluster_bootstrap_geotax()` — fit
   logistic regressions of sharing probability on phylogenetic distance,
   with parasite-level cluster bootstrap uncertainty and formal
   between-clade slope comparison.

Legacy single-focal-host functions (`fit_geotax_model()`,
`bootstrap_geotax_model()`, and the deprecated `get_log_reg_coefficients()`
family) are kept for backward compatibility.

## Example: taxonomic tree

Compute a taxonomic tree from a table with taxonomic ranks (bark beetle
hosts from the original paper):

```{r example}
library(geotax)
library(dplyr) ## this is for clean the data if need it
data("tax_table")
clean_tree <- distinct(tax_table)

# clean the data
taxonomic_tree <- geotax::get_taxonomical_tree(clean_tree, power = 1)
plot(taxonomic_tree, type = "radial", show.tip.label = FALSE)
ape::cophenetic.phylo(taxonomic_tree)[1:5, 1:5]
```

## Example: pairwise host-sharing model

```{r pairwise}
pairs <- prepare_pair_data(beetleTreeInteractions, phy_dist)
boot <- cluster_bootstrap_geotax(pairs, n_boot = 100, seed = 42)
boot
```

## Example: comparing clades

See the vignette `vignette("caligus-vs-lepeophtheirus")` for a full case
study comparing sea lice genera *Caligus* (generalist) and
*Lepeophtheirus* (specialist) on marine fishes, using interaction records
from the [cofid](https://github.com/alrobles/cofid) package and the Fish
Tree of Life phylogeny.
