---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geotax

<!-- badges: start -->
<!-- badges: end -->

This packages aims to model the probability that two host share a pathogen
given a phylogenetic (or taxonomic) distance. Please link to the original
[paper](https://doi.org/10.3389/fams.2017.00017)



## Installation

You can install the development version of geotax from [GitHub](https://github.com/alrobles/geotax) with:

``` r
# install.packages("devtools")
devtools::install_github("alrobles/geotax")
```

## Examples

geotax comes with basic tools. For example compute a taxonomic tree from a 
table with taxonomic ranks. In this example we load a taxonomic tree of species
that has been record with bark beetles. We use the table provide in the paper.

We can also plot this tree  and explore the distance matrix with the help of ape package


```{r example}
library(geotax)
library(dplyr) ## this is for clean the data if need it
data("tax_table")
clean_tree <- distinct(tax_table)

# clean the data
taxonomic_tree <- geotax::get_taxonomical_tree(clean_tree, power = 1)
plot(taxonomic_tree, type = "radial", show.tip.label = FALSE)
ape::cophenetic.phylo(taxonomic_tree)[1:5, 1:5]
```

