---
title: "Comparing clades: Caligus vs Lepeophtheirus sea lice on marine fishes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing clades: Caligus vs Lepeophtheirus sea lice on marine fishes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
has_cofid <- requireNamespace("cofid", quietly = TRUE)
```

This vignette shows the full geotax v2 workflow on a real host-parasite
system: sea lice (Copepoda: Caligidae) parasitizing marine fishes. We ask
whether the probability that two fish species share a sea louse decays
faster with host phylogenetic distance in *Lepeophtheirus* than in
*Caligus* — i.e., whether *Lepeophtheirus* is a phylogenetic specialist
and *Caligus* a generalist.

Interaction records come from the
[cofid](https://github.com/alrobles/cofid) package (copepod-fish
interaction database) and the host phylogeny from the
[Fish Tree of Life](https://fishtreeoflife.org/) (Rabosky et al. 2018),
pruned to the recorded hosts and shipped as the `fish_tree` data object.

```{r setup, eval = has_cofid}
library(geotax)
library(ape)
```

## 1. Interaction data

We keep one record per copepod species / fish species pair and derive
the copepod genus, which is the clade-level grouping we want to compare.

```{r data, eval = has_cofid}
interactions <- cofid::cofid
interactions$genus <- sub(" .*", "", interactions$source_taxon_name)
interactions <- interactions[
  interactions$genus %in% c("Caligus", "Lepeophtheirus"),
  c("source_taxon_name", "target_taxon_name", "genus")
]
table(unique(interactions[, c("source_taxon_name", "genus")])$genus)
```

## 2. Host phylogenetic distances

`fish_tree` is already pruned to the recorded hosts present in the Fish
Tree of Life; its cophenetic matrix gives the phylogenetic distance
between every pair of hosts.

```{r phydist, eval = has_cofid}
data(fish_tree)
phydist <- cophenetic(fish_tree)
dim(phydist)
```

## 3. Pairwise host-sharing data

`prepare_pair_data()` expands the interaction table into all
(focal host, target host) pairs per parasite. Every known host is used
as a focal host — unlike the legacy workflow, which sampled a single
focal host per parasite — so the data set is deterministic and uses all
available information.

```{r pairs, eval = has_cofid}
pairs <- prepare_pair_data(interactions, phydist, group = "genus")
nrow(pairs)
head(pairs)
```

## 4. Comparing slopes between clades

`compare_geotax_slopes()` fits a single logistic regression with a
genus-by-distance interaction and quantifies uncertainty with a
parasite-level cluster bootstrap, which respects the non-independence
of pairs that share a parasite.

```{r compare, eval = has_cofid}
cmp <- compare_geotax_slopes(pairs, n_boot = 20, seed = 42)
cmp
```

A more negative slope means the probability of sharing a parasite
decays faster with phylogenetic distance. The *Lepeophtheirus* slope is
steeper than the *Caligus* slope and the interaction term is negative,
supporting the interpretation of *Lepeophtheirus* as a phylogenetic
specialist and *Caligus* as a generalist.

## 5. Predicted sharing curves

```{r curves, eval = has_cofid}
cf <- coef(cmp$fitted_model)
d <- seq(0, max(pairs$phydist), length.out = 200)
p_cal <- plogis(cf["(Intercept)"] + cf["phydist"] * d)
p_lep <- plogis(cf["(Intercept)"] + cf["genusLepeophtheirus"] +
                  (cf["phydist"] + cf["phydist:genusLepeophtheirus"]) * d)
plot(d, p_cal, type = "l", col = "#1b9e77", lwd = 2, ylim = c(0, max(p_cal)),
     xlab = "Host phylogenetic distance",
     ylab = "P(sharing a sea louse)")
lines(d, p_lep, col = "#d95f02", lwd = 2)
legend("topright", legend = c("Caligus (generalist)",
                              "Lepeophtheirus (specialist)"),
       col = c("#1b9e77", "#d95f02"), lwd = 2, bty = "n")
```

## References

- Robles-Fernández, Á. L. & Lira-Noriega, A. (2017). Combining
  phylogenetic and occurrence information for risk assessment of pest
  and pathogen interactions with host plants. *Frontiers in Applied
  Mathematics and Statistics*, 3:17.
  [doi:10.3389/fams.2017.00017](https://doi.org/10.3389/fams.2017.00017)
- Rabosky, D. L. et al. (2018). An inverse latitudinal gradient in
  speciation rate for marine fishes. *Nature*, 559, 392-395.
- Tinoco-Domínguez, E., Amancio, G., Robles-Fernández, Á. L. &
  Lira-Noriega, A. (2025). Interaction network of *Phoradendron* and its
  hosts. *American Journal of Botany*, e70025.
